---
title: 'Mutate: Creation Function Demonstration'
output:
  html_document:
    css: "styles.css"
    toc: true
    toc_float: true
    theme: cosmo
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #a6873b;
    color: black;
    font-weight: bold;
}
</style>
```
```{=html}
<style>
  .nav-pills>li>a:hover, .nav-pills>li>a:focus, .nav-pills>li.active>a,     .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus{
     background-color: #a6873b;
     font-weight: bold;
          color: black;
     }
</style>
```
```{=html}
<style>
  .nav-pills>li>a {
     color: black;
     }
</style>
```
```{=html}
<style>
  a {
     color: #a6873b;
     text-decoration: none;
     }
</style>
```
```{=html}
<style>
  a:hover {
     color: #a6873b;
     font-weight: bold;
     }
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages

First, let's load all relevant packages and explore the dataset

```{r warning = FALSE, message = FALSE}
require(tidyverse)
require(palmerpenguins)
require(DT)
```

I'm going to select out several columns for ease of data presentation.  Note that throughout the code below, the
`%>% datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))` syntax is
used to print the data frame.

```{r}
penguins <- select(penguins, -c(bill_depth_mm, year, flipper_length_mm, island))
penguins %>% 
  datatable(options = list(pageLength = 344, scrollX = 250, scrollY = 300, dom = "tpi"))
```

# Arithmetic Operations

## Basic Arithmetic

First, let's look at some basic arithmetic operations, i.e., `+`, `-`,
`*`, and `/`.

We can simply apply these functions to a single variable, like this:

```{r}
mutate(penguins, body_mass_2 = body_mass_g * 2) %>% 
  datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))
```

<br>

Or we can apply these to multiple variables to create a new variable,
like this:

```{r}
mutate(penguins, body_mass_to_bill_length = body_mass_g / bill_length_mm) %>% 
  datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))
```

<br>

## Aggregate Arithmetic

We can also use these basic arithmetic operations along with aggregate
arithmetic operations, such as `mean` or `median`.

The below example centers body mass at the mean, so all positive values
are above average and all negative values are below average. Note that
we had to use `na.rm = TRUE` because there are missing body mass values
in the data set.

```{r}
mutate(penguins, body_mass_centered = body_mass_g - mean(body_mass_g, na.rm = TRUE)) %>% 
  datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))
```

<br>

We could use two aggregate arithmetic functions in a `mutate`, but the
result is unlikely to be what we want.

```{r}
mutate(penguins, bad_var = mean(body_mass_g, na.rm = TRUE) - mean(bill_length_mm, na.rm = TRUE)) %>% 
  datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))
```

<br>

## Log Transformation

Log transformation can be easily performed using the `log` functions.
`log` will perform a natural log transformation, `log2` will perform a
log transformation with base 2, and `log10` will perform a log
transformation with base 10. You can specifify other bases using the
`base` argument within the `log` function (e.g., `log(x, base = 4)`,
where x is a vector of numbers).

```{r}
mutate(penguins, log_body_mass = log(body_mass_g)) %>% 
  datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))
```

<br>

## Modular Arithmetic

Modular arithmetic is helpful in breaking integers into pieces.
**Integer division** yields the full integer from the arithmetic
operation and **remainder** returns just the remainder from the
operation.

```{r}
mutate(penguins, 
       body_mass_div3_int = body_mass_g %/% 3,
       body_mass_div3_rem = body_mass_g %% 3) %>% 
  datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))
```

<br>

# Offsets

Offsets are very useful in creating new variables, especially when you
have multiple observations on some grouped factor (e.g., multiple
observations per person in longitudinal data). `lag` compares data to a
previous instance (behind the current value) and `lead` compares data in
a next instance (ahead of the current value). You can use the `n`
argument within these functions to specify the distance between compared
values

## Lag

```{r}
mutate(penguins, body_mass_lag = body_mass_g - lag(body_mass_g, n = 1)) %>% 
  datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))
```

## Lead

```{r}
mutate(penguins, body_mass_lag = body_mass_g - lead(body_mass_g, n = 2)) %>% 
  datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))
```

# Cumulative Aggregates

Cumulative aggreative functions allow for arithmetic calculations over
cumulative sets of numbers. The cumulative aggreate functions include
`cumsum`, `cumprod`, `cummin`, `cummmax`, and `cummean`.

Here's a demonstration of the `cumsum` function.

```{r}
mutate(penguins, body_mass_cumsum = cumsum(body_mass_g)) %>% 
  datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))
```

Note how NA values are produced when data are missing. Let's exlcude all
missing cases and run this function again.

```{r}
penguins_full <- filter(penguins, !is.na(body_mass_g))
mutate(penguins_full, body_mass_cumsum = cumsum(body_mass_g)) %>% 
  datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))
```

<br>

Here's a demonstration of the `cummin` function.

```{r}
mutate(penguins_full, body_mass_cummin = cummin(body_mass_g)) %>% 
  datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))
```

And last let's look at `cummean`

```{r}
mutate(penguins_full, body_mass_cummean = cummean(body_mass_g)) %>% 
  datatable(options = list(scrollX = 250, scrollY = 300, dom = "tpi"))
```

# Ranking

There are several functions available to compute rank functions in R:
<br>

1.  `min_rank(x)`: ranks from smallest to largest <br>
2.  `min_rank(desc(x))`: ranks from largest to smallest <br>
3.  `dense_rank()`: like `min_rank`, but with no gaps between ranks <br>
4.  `percent_rank()`: a number between 0 and 1 computed by rescaling
    min_rank to [0, 1] <br>
5.  `cume_dist()`: a cumulative distribution function. Proportion of all
    values less than or equal to the current rank. <br>
6.  `ntile()`: a rough rank, which breaks the input vector into n
    buckets. The size of the buckets may differ by up to one, larger
    buckets have lower rank. <br>

<br>

In the sample below, I use the `%>%` operator to chain together a string
a functions. We'll talk about this next week, but basically this
operator takes the result of the prior line and sends it through to the
next line, where it becomes the first argument in the function. I've
arranged and selected specific columns for ease of presentation.

<br>

```{r}
mutate(penguins, 
       min = min_rank(body_mass_g), 
       max = min_rank(desc(body_mass_g)), 
       dense = dense_rank(body_mass_g), 
       cumulative = cume_dist(body_mass_g), 
       percent = percent_rank(body_mass_g), 
       n10 = ntile(body_mass_g, n = 10)) %>% 
  arrange(min) %>% 
  select(body_mass_g, min, max, dense, cumulative, percent, n10) %>% 
  datatable(options = list(pageLength = 100, scrollX = 250, scrollY = 300, dom = "tpi"))
```


# Logical Comparisons

Logical comparisons are often performed to re-code variables using the `case_when` function.

```{r}
mutate(penguins, high_body_mass = 
         case_when(body_mass_g > 4000 ~ 1, 
                   body_mass_g <= 4000 ~ 0)) %>% 
  datatable(options = list(pageLength = 10, scrollX = 250, scrollY = 300, dom = "tpi"))
```

