---
title: "Data Visualization I:  Solutions"
output:
  html_document:
    css: "styles.css"
    theme: cosmo
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #8a2377;
    color: white;
    font-weight: bold;
}
</style>
```
```{=html}
<style>
  .nav-pills>li>a:hover, .nav-pills>li>a:focus, .nav-pills>li.active>a,     .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus{
     background-color: #8a2377;
     font-weight: bold;
          color: black;
     }
</style>
```
```{=html}
<style>
  .nav-pills>li>a {
     color: black;
     }
</style>
```
```{=html}
<style>
  a {
     color: #8198b9;
     text-decoration: none;
     }
</style>
```
```{=html}
<style>
  a:hover {
     color: #8198b9;
     font-weight: bold;
     }
</style>
```
```{r}
require(tidyverse)
```

1.  Run `ggplot(data = mpg)`. What do you see? <br>

```{r}
ggplot(data = mpg)
```

This code creates an empty plot. The ggplot() function creates the
background of the plot, but since no layers were specified with geom
function, nothing is drawn. <br> <br> 2. Make a scatterplot of hwy
versus cyl using the mpg dataset. <br>

```{r}
ggplot(mpg, aes(x = cyl, y = hwy)) +
  geom_point()
```

<br> <br> 3. What happens if you make a scatterplot of class versus drv?
Why is the plot not useful? <br>

```{r}
ggplot(mpg, aes(x = class, y = drv)) +
  geom_point()
```

<br> A scatter plot is not a useful display of these variables since
both drv and class are categorical variables. Since categorical
variables typically take a small number of values, there are a limited
number of unique combinations of (x, y) values that can be displayed. In
this data, drv takes 3 values and class takes 7 values, meaning that
there are only 21 values that could be plotted on a scatterplot of drv
vs. class. In this data, there 12 values of (drv, class) are observed.

```{r}
count(mpg, drv, class)
```

A simple scatter plot does not show how many observations there are for
each (x, y) value. As such, scatterplots work best for plotting a
continuous x and a continuous y variable, and when all (x, y) values are
unique. <br>

Below are two alternative plots that are well-suited for displaying
categorical x categorical visualizations.

```{r}
ggplot(mpg, aes(x = class, y = drv)) +
  geom_count()
```

```{r}
mpg %>%
  count(class, drv) %>%
  ggplot(aes(x = class, y = drv)) +
    geom_tile(mapping = aes(fill = n))
```

<br> <br> 4. Which variables in mpg are categorical? Which variables are
continuous (Hint: type ?mpg to read the documentation for the dataset).
How can you see this information when you run mpg? <br>

The following list contains the categorical variables in mpg:

-   manufacturer

-   model

-   trans

-   drv

-   fl

-   class

The following list contains the continuous variables in mpg:

-   displ

-   year

-   cyl

-   cty

-   hwy

In the printed tibble, angled brackets at the top of each column provide
type of each variable.

```{r}
mpg
```

Those with `<chr>` above their columns are categorical, while those with
`<dbl>` or `<int>` are continuous.

We could also use the `str` function to evaluate object type

```{r}
str(mpg)
```

Finally, `glimpse` also produces an `str`-like summary of a data frame.

```{r}
glimpse(mpg)
```

<br> <br> 5. What happens if you map an aesthetic to something other
than a variable name, like `aes(color = displ < 5)` 6. What happens if
you facet on a continuous variable? <br>

```{r}
ggplot(mpg, aes(x = displ, y = hwy, colour = displ < 5)) +
  geom_point()
```

<br>

Aesthetics can also be mapped to expressions like `displ < 5`. The
`ggplot()` function behaves as if a temporary variable was added to the
data with values equal to the result of the expression. In this case,
the result of `displ < 5` is a logical variable which takes values of
`TRUE` or `FALSE`.

<br> <br> 6. What happens if you facet on a continuous variable?

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(. ~ cty)
```

<br>

The continuous variable is converted to a categorical variable, and the
plot contains a facet for each distinct value.

<br> <br> 7. What do the empty cells in a plot with
`facet_grid(drv~cyl)` mean? How do they relate to this plot? <br>
`ggplot(data = mpg) + geom_point(aes(x = drv, y = cyl)) <br>`

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = hwy, y = cty)) +
  facet_grid(drv ~ cyl)
```

<br>

The empty cells (facets) in this plot are combinations of `drv` and
`cyl` that have no observations. These are the same locations in the
scatter plot of `drv` and `cyl` that have no points.

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = cyl, y = drv)) + 
  scale_y_discrete(limits = rev)
```

<br> <br> 8. What geom would you use to draw a line chart? A boxplot? A
histogram? An area chart? <br>

-   line chart: `geom_line()`

-   boxplot: `geom_boxplot()`

-   histogram: `geom_histogram()`

-   area chart: `geom_area()`

<br> <br> 9. What does `show.legend = FALSE` do? What happens if you
remove it?

The theme option `show.legend = FALSE` hides the legend box.

Consider this example earlier in the chapter.

```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, colour = drv),
    show.legend = FALSE
  )
```

<br>

In that plot, there is no legend. Removing the `show.legend` argument or
setting `show.legend = TRUE` will result in the plot having a legend
displaying the mapping between colors and `drv`.

```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, colour = drv))
```

<br> <br> 10. What does `geom_col` do? How is it different from
`geom_bar`? <br>

The `geom_col()` function has different default stat than `geom_bar()`.
The default stat of `geom_col()` is `stat_identity()`, which leaves the
data as is. The `geom_col()` function expects that the data contains `x`
values and `y` values which represent the bar height.

The default stat of `geom_bar()` is `stat_count()`. The `geom_bar()`
function only expects an `x` variable. The stat, `stat_count()`,
preprocesses input data by counting the number of observations for each
value of `x`. The `y` aesthetic uses the values of these counts.

<br> <br> 11. What is the problem with this plot? How could you improve
it?

```{r echo = FALSE, warning = FALSE, message = FALSE}
require(tidyverse)
```

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point()
```

<br>

There is overplotting because there are multiple observations for each
combination of `cty` and `hwy` values.

I would improve the plot by using a jitter position adjustment to
decrease overplotting.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point(position = "jitter")
```

<br> <br> 13. What parameters to `geom_jitter` control the amount of
jittering? <br>

From the
[`geom_jitter()`](https://ggplot2.tidyverse.org/reference/geom_jitter.html)
documentation, there are two arguments to jitter:

-   `width` controls the amount of horizontal displacement, and

-   `height` controls the amount of vertical displacement.

The defaults values of `width` and `height` will introduce noise in both
directions. Here is what the plot looks like with the default values of
`height` and `width`.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point(position = position_jitter())
```

<br>

However, we can change these parameters. Here are few a examples to
understand how these parameters affect the amount of jittering.
When`width = 0` there is no horizontal jitter.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter(width = 0)
```

<br>

When `width = 20`, there is too much horizontal jitter.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter(width = 20)
```

<br>

When `height = 0`, there is no vertical jitter.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter(height = 0)
```

When `height = 15`, there is too much vertical jitter.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter(height = 15)
```

<br>

When `width = 0` and `height = 0`, there is neither horizontal or
vertical jitter, and the plot produced is identical to the one produced
with `geom_point()`

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter(height = 0, width = 0)
```

Note that the `height` and `width` arguments are in the units of the
data. Thus `height = 1` (`width = 1`) corresponds to different relative
amounts of jittering depending on the scale of the `y` (`x`) variable.
The default values of `height` and `width` are defined to be 80% of the
`resolution()` of the data, which is the smallest non-zero distance
between adjacent values of a variable. When `x` and `y` are discrete
variables, their resolutions are both equal to 1, and `height = 0.4` and
`width = 0.4` since the jitter moves points in both positive and
negative directions.

The default values of `height` and `width` in `geom_jitter()` are
non-zero, so unless both `height` and `width` are explicitly set set 0,
there will be some jitter.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter()
```
