---
title: "Data Visualization I:  Problems"
output:
  html_document:
    css: "styles.css"
    theme: cosmo
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #a6873b;
    color: black;
    font-weight: bold;
}
</style>
```
```{=html}
<style>
  .nav-pills>li>a:hover, .nav-pills>li>a:focus, .nav-pills>li.active>a,     .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus{
     background-color: #a6873b;
     font-weight: bold;
          color: black;
     }
</style>
```
```{=html}
<style>
  .nav-pills>li>a {
     color: black;
     }
</style>
```
```{=html}
<style>
  a {
     color: #a6873b;
     text-decoration: none;
     }
</style>
```
```{=html}
<style>
  a:hover {
     color: #a6873b;
     font-weight: bold;
     }
</style>
```





1.  Run `ggplot(data = mpg)`.  What do you see? <br> <br> 
2.  Make a scatterplot of hwy versus cyl using the mpg dataset. <br> <br> 
3.  What happens if you make a scatterplot of class versus drv?  Why is the plot not useful? <br> <br> 
4.  Which variables in pg are categorical?  Which variables are continuous (Hint:  type ?mpg to read the documentation for the dataset).  How can you see this information when you run mpg? <br> <br> 
5.  What happens if you map an aesthetic to something other than a variable name, like `aes(color = displ < 5)` <br> <br> 
6.  What happens if you facet on a continuous variable? <br> <br> 
7.  What do the empy cells in a plot with `facet_grid(drv~cyl)` mean?  How do they relate to this plot? <br>
`ggplot(data = mpg) + geom_point(aes(x = drv, y = cyl))` <br> <br> 
8.  What geom would you use to draw a line chart?  A boxplot?  A histogram?  An area chart? <br> <br> 
9.  What does `show.legend = FALSE` do?  What happens if you remove it?   <br> <br> 
10.  What does `geom_col` do?  How is it different from `geom_bar`? <br> <br> 
11.  What is the problem with this plot?  How could you improve it?
```{r echo = FALSE, warning = FALSE, message = FALSE}
require(tidyverse)
```

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point()
```
<br>  <br> 
12.  What parameters to `geom_jitter` control the amount of jittering? <br> <br> 
