---
title: "Tidy Data Project:  Problems"
output:
  html_document:
    css: "styles.css"
    theme: cosmo
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #a6873b;
    color: black;
    font-weight: bold;
}
</style>
```
```{=html}
<style>
  .nav-pills>li>a:hover, .nav-pills>li>a:focus, .nav-pills>li.active>a,     .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus{
     background-color: #a6873b;
     font-weight: bold;
          color: black;
     }
</style>
```
```{=html}
<style>
  .nav-pills>li>a {
     color: black;
     }
</style>
```
```{=html}
<style>
  a {
     color: #a6873b;
     text-decoration: none;
     }
</style>
```
```{=html}
<style>
  a:hover {
     color: #a6873b;
     font-weight: bold;
     }
</style>
```
For this exercise, you'll continue to explore the WHO datasets that we
looked at in module 9. Load the `tidyverse` package. This code is
reproduced from the chapter because it is needed by the exercises.

```{r}
tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
```

<br> <br>

1.  Recreate the plot showing change in cases over time using table2
    instead of table1. What do you need to do first? <br>

Before creating the plot with change in cases over time, we need to
filter table to only include rows representing cases of TB.

```{r}
table2 %>%
  filter(type == "cases") %>%
  ggplot(aes(year, count)) +
  geom_line(aes(group = country), colour = "grey50") +
  geom_point(aes(colour = country)) +
  scale_x_continuous(breaks = unique(table2$year)) +
  ylab("cases")
```

<br> <br>

2.  Why does this code fail?

```{r error = TRUE}
table4a %>% 
  pivot_longer(c(1999, 2000), names_to = "year", values_to = "cases")
```

<br> <br>

3.  

    3.  What would happen if you widen this table? Why? How could you
        add a new column to uniquely identify each value?

```{r}
people <- tribble(
  ~name, ~key, ~value,
  #-----------------|--------|------
  "Phillip Woods",  "age", 45,
  "Phillip Woods", "height", 186,
  "Phillip Woods", "age", 50,
  "Jessica Cordero", "age", 37,
  "Jessica Cordero", "height", 156
)
glimpse(people)
```

<br> <br>

4.  Tidy the simple tibble below. Do you need to make it wider or
    longer? What are the variables?

```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes", NA, 10,
  "no", 20, 12
)
```

<br> 
<br>

5.  What do the extra and fill arguments do in separate()? Experiment with the various options for the following two toy datasets. 

```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"))
```

```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"))
```

<br>
<br>

6.  Both `unite()` and `separate()` have a remove argument. What does it do? Why would you set it to FALSE?

<br>
<br>

7.  What does the direction argument to fill() do?