---
title: "Data Transformation:  Problems"
output:
  html_document:
    css: "styles.css"
    theme: cosmo
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #a6873b;
    color: black;
    font-weight: bold;
}
</style>
```
```{=html}
<style>
  .nav-pills>li>a:hover, .nav-pills>li>a:focus, .nav-pills>li.active>a,     .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus{
     background-color: #a6873b;
     font-weight: bold;
          color: black;
     }
</style>
```
```{=html}
<style>
  .nav-pills>li>a {
     color: black;
     }
</style>
```
```{=html}
<style>
  a {
     color: #a6873b;
     text-decoration: none;
     }
</style>
```
```{=html}
<style>
  a:hover {
     color: #a6873b;
     font-weight: bold;
     }
</style>
```




For this project, you'll be working with the flights data set in the nycflights13 package.  If you haven't already done so, install the nycflights13 package (`install.packages("nycflights13")`) and load it (`require(nycflights13)`) along with the tidyverse (`require(tidyverse)`).

```{r warning = FALSE, message = FALSE}
require(tidyverse)
require(nycflights13)
```


To get you oriented, let's first take a look at the flights data set

```{r}
flights
```


```{r}
glimpse(flights)
```

Consult the help documentation (`?flights`) for more information on the variables in this data set.

# Filter 

1. Find all flights that
   a. Had an arrival delay of two or more hours
   b. Flew to Houston
   c. Were operated by United, American, or Delta
   d. Departed in summer (July, August, and September)
   e. Arrived more tehan two hours late, but didn't leave
   f. Were delayed by at least one hour, but made up over 30 minutes in flight
   g. Departed between midnight and 6 a.m. (inclusive) <br> <br> 
2. How many flights have a missing `dep_time`?  What other variables are missing?  What might these rows represent? <br> <br> 

# Arrange

3.  How could you use `arrange()` to sort all missing values to the start? (Hint:  use `is.na()`) <br> <br> 
4.  Sort flights to find the most delayed flights.  Find the flights that left earliest. <br> <br> 
5.  Sort flights to find the fastest flights. <br> <br> 

# Select

6. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from `flights`. <br> <br> 
7. What happens if you include the name of a variable multiple times in a select() call? <br> <br> 

# Mutate

8. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight. <br> <br> 
9. Compare air_time with arr_time - dep_time. What do you expect to see? What do you see? What do you need to do to fix it? <br> <br> 

# Summarize

10.  Use the following code to identify canceled flights: `is.na(arr_delay) | is.na(dep_delay)`.  Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay? <br> <br> 
